Optimal synaptic strategies for different timescales of memory

### 2018-03-19 Paper/synapse_title.tex
separated main, sup, etc

\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
% Use the lineno option to display guide line numbers if required.
% Note that the use of elements such as single-column equations
% may affect the guide line number alignment.

\templatetype{pnasresearcharticle} % Choose template
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} = Template for a one-column mathematics article
% {pnasinvited} = Template for a PNAS invited submission

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% ADDITIONAL OPTIONAL STYLE FILES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{graphicx}
%\usepackage{graphics} for finding documentation only
\pdfsuppresswarningpagegroup=1
\usepackage{epstopdf}
\epstopdfsetup{update,suffix=-generated}
%
\IfFileExists{"C:/Progra\string~1/Inkscape/inkscape.exe"}{\newcommand{\inkscapedir}{C:/Progra\string~1/Inkscape/}}{}
\IfFileExists{"/Applications/Inkscape.app/Contents/Resources/bin/inkscape"}{\newcommand{\inkscapedir}{/Applications/Inkscape.app/Contents/Resources/bin/}}{}
\providecommand{\inkscapedir}{}
\epstopdfDeclareGraphicsRule{.svg}{pdf}{.pdf}{%
  "\inkscapedir inkscape" -f #1 -C -A \OutputFile}
%
\usepackage{adjustbox}
\newcommand{\aligntop}[1]{\adjustbox{valign=t}{#1}}
%\usepackage[inline]{enumitem}
\newlist{myenuma}{enumerate*}{4}
\setlist[myenuma]{label=(\alph*),ref=\alph*}
\newcommand{\panel}[1]{\item\aligntop{#1}}
\graphicspath{{"Figs/"}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amssymb,amsfonts}
\usepackage{amsthm}
%\usepackage[notref,notcite]{showkeys}
%\usepackage[centertags]{amsmath}
%\usepackage{bibunits}
%\usepackage{hyperref}
\hypersetup{bookmarksopenlevel=1,bookmarksnumbered}
\usepackage[capitalise,nameinlink]{cleveref}
%\Crefname{appendix}{App.}{Apps.}
\creflabelformat{equation}{[#2#1#3]}
\crefrangelabelformat{equation}{[#3#1#4--#5#2#6]}
%\crefformat{equation}{#2Eq.~[#1]#3}
%\crefrangeformat{equation}{Eqs.~[#3#1#4--#5#2#6]}
%\crefmultiformat{equation}{Eqs.~[#2#1#3}{,2#1#3]}{,#2#1#3}{,#2#1#3]}
%\crefrangemultiformat{equation}{Eqs.~[#3#1#4--#5#2#6}{,#3#1#4--#5#2#6]}{,#3#1#4--#5#2#6}{,#3#1#4--#5#2#6]}
\newcommand{\main}{the main paper}
\newcommand{\supp}{supporting information}
% ----------------------------------------------------------------
\newtheoremstyle{sldefinition}%
  {3pt}%space above
  {3pt}%space below
  {}%body font
  {}%indent amount
  {\bfseries}%theorem head font
  {}%theorem head punctuation
  {\newline}%space after head
  {\thmname{#1}\thmnumber{ #2}:{\bfseries\thmnote{ #3}}}%head spec
\newtheoremstyle{slplain}%
  {3pt}%space above
  {3pt}%space below
  {}%body font
  {}%indent amount
  {\bfseries}%theorem head font
  {}%theorem head punctuation
  {\newline}%space after head
  {\thmname{#1}\thmnumber{ #2}{\bfseries\thmnote{ (#3)}}: }%head spec
%
\theoremstyle{slplain}
\newtheorem{thm}{Theorem}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{lem}[thm]{Lemma}
%
\theoremstyle{sldefinition}
\newtheorem{defn}{Definition}
%
\theoremstyle{remark}
\newtheorem*{notn}{Notation}
\newtheorem*{rem}{Remark}
% ----------------------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% OPTIONAL MACRO DEFINITIONS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\ra}{\rightarrow}
\newcommand{\lr}{\leftrightarrow}
\newcommand{\cdt}{\!\cdot\!}
\newcommand{\vp}{\vspace{0.5cm}}
\newcommand{\hp}{\hspace{0.5cm}}
%
%e.g., i.e. with normal spaces
\newcommand{\eg}{e.g.\ }
\newcommand{\ie}{i.e.\ }
\newcommand{\cf}{cf.\ }
\newcommand{\etc}{etc.\ }
\newcommand{\wrt}{wrt.\ }
\newcommand{\Eg}{E.g.\ }
\newcommand{\Ie}{I.e.\ }
\newcommand{\Cf}{Cf.\ }
\newcommand{\Etc}{Etc.\ }
%
\usepackage{delim}
% brackets etc.
\delimdef\prn#1{\dleft( #1 \dright)}
\delimdef\brc#1{\dleft\{ #1 \dright\}}
\delimdef\brk#1{\dleft[ #1 \dright]}
\delimdef\abs#1{\dleft\lvert #1 \dright\rvert}
\delimdef\nrm#1{\dleft\lVert #1 \dright\rVert}
\delimdef\av#1{\dleft\langle #1 \dright\rangle}
\delimdef\avc#1#2{\dleft\langle #1 \dmiddle\vert #2 \dright\rangle}
%\newenvironment{cases}{\dleft\{\begin{aligned}}{\end{aligned}\dright.}
%
% Sets
\delimdef\set#1#2{\dleft\{ #1 \dmiddle\vert #2 \dright\}}
\delimdef\cond#1#2{\!\dleft( #1 \dmiddle\vert #2 \dright)}
\delimdef\condb#1#2{\!\dleft[ #1 \dmiddle\vert #2 \dright]}
%
% Derivatives, etc. First argument is optional.
\newcommand{\diff}[3][]{\frac{\mathrm{d}^{#1} #2}{\mathrm{d}{#3}^{#1}}}
\newcommand{\pdiff}[3][]{\frac{\partial^{#1} #2}{\partial {#3}^{#1}}}
\newcommand{\pdiffc}[3][]{\left(\frac{\partial #2}{\partial {#3}}\right)_{\!\!#1}}
\newcommand{\intd}[2][]{\int #1\!\!\dr #2\,}
%
% Un-italicised letters
\newcommand{\dr}{\mathrm{d}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\ir}{\mathrm{i}}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\Det}{Det}
\DeclareMathOperator{\cov}{Cov}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\bias}{Bias}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\sech}{sech}
\DeclareMathOperator{\diag}{diag}
%
% The default \Im and \Re look crap
\renewcommand{\Im}{\operatorname{\mathfrak{Im}}}
\renewcommand{\Re}{\operatorname{\mathfrak{Re}}}
%
% logic
\newcommand{\means}{\Longleftrightarrow}
\newcommand{\requires}{\Longleftarrow}
%
%additional symbols:
\newcommand{\half}{\frac{1}{2}}
\newcommand{\CO}{\mathcal{O}}
\newcommand{\CL}{\mathcal{L}}
\newcommand{\dt}{\dr t}
%
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\newcommand{\wv}{\vec{w}}
\newcommand{\wvi}{\vec{w}_\text{ideal}}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distributions
\newcommand{\prob}{\mathbf{p}}
\newcommand{\eq}{\boldsymbol{\pi}}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols for matrices
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\arow}{\boldsymbol{\xi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{q}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%sets
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
%eigenmodes
\newcommand{\uv}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\CI}{\mathcal{I}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Subhaneil Lahiri}
\author[a]{Surya Ganguli}

\affil[a]{Department of Applied Physics, Stanford University, Stanford CA}

@@

% Please add here a significance statement to explain the relevance of your work
\significancestatement{Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Please provide details of author contributions here.}
\authordeclaration{Please declare any conflict of interest here.}
%\equalauthors{\textsuperscript{1}A.O.(Author One) and A.T. (Author Two) contributed equally to this work (remove if not applicable).}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: sulahiri@stanford.edu}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{synaptic plasticity | learning | memory | neuroscience}

%\abbreviations{SNR, signal-to-noise ratio; LTP, long term potentiation;LTD, long term depression}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\begin{article}
\begin{abstract}
  An incredible gulf separates theoretical models of synapses, often described solely by a single scalar value denoting the size of a postsynaptic potential, from the immense complexity of molecular signaling pathways underlying real synapses.
  To understand the functional contribution of such molecular complexity to learning and memory, it is essential to expand our theoretical conception of a synapse from a single scalar to an entire dynamical system with many internal molecular functional states.
  Moreover, theoretical considerations alone demand such an expansion; network models with scalar synapses assuming finite numbers of distinguishable synaptic strengths have strikingly limited memory capacity.
  This raises the fundamental question, how does synaptic complexity give rise to memory?
  To address this, we develop new mathematical theorems elucidating the relationship between the structural organization and memory properties of complex synapses that are themselves molecular networks.
  Moreover, in proving such theorems, we uncover a framework, based on first passage time theory, to impose an order on the internal states of complex synaptic models, thereby simplifying the relationship between synaptic structure and function.

  Overall, we uncover general design principles governing the functional organization of complex molecular networks, and suggest new experimental observables in synaptic physiology, based on first %passage time theory, that connect molecular complexity to memory.
\end{abstract}

@@

\setboolean{displaywatermark}{false}
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{article}

% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
% You should only change this length when you've finalised the article contents.
\verticaladjustment{-2pt}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

\dropcap{I}t is widely thought that our very ability to remember the past over long time scales depends crucially on our ability to modify synapses in our brain in an experience dependent manner.
Classical models of synaptic plasticity model synaptic efficacy as an analog scalar value, denoting the size of a post-synaptic potential injected into one neuron from another.
Theoretical work has shown that such models have a reasonable, extensive memory capacity, in which the number of long term associations that can be stored by a neuron is proportional its number of afferent synapses \cite{Hopfield1982model,Amit1985hopfield,Gardner1988perceptron}.
However, recent experimental work has shown that many synapses are more digital than analog; they cannot robustly assume an infinite continuum of analog values, but rather can only take on a finite number of distinguishable strengths, a number than can be as small as two \cite{Bliss1993LTP,Petersen1998allornone,O'Connor2005switch} \cite[though see][] {Enoki2009graded}.
This one simple modification leads to a catastrophe in memory capacity:  classical models with digital synapses, when operating in a palimpsest mode in which the ongoing storage of new memories can overwrite previous memories, have a memory capacity proportional to the logarithm of the number of synapses \cite{amit1992constraints,amit1994learning}.
Intuitively, when synapses are digital, the storage of a new memory can flip a population of synaptic switches, thereby rapidly erasing previous memories stored in the same synaptic population. This result indicates that the dominant theoretical basis for the storage of long term memories in modifiable synaptic switches is flawed.

Recent work \cite{Fusi2005cascade,Fusi2007multistate,Leibold2008serial} has suggested that a way out of this logarithmic catastrophe is to expand our theoretical conception of a synapse from a single scalar value to an entire stochastic dynamical system in its own right.
This conceptual expansion is further necessitated by the experimental reality that synapses contain within them immensely complex molecular signaling pathways, with many internal molecular functional states \cite[\eg see][]{Bliss1993LTP,Bredt2003AMPA,Coba2009phosphorylation}.
While externally, synaptic efficacy could be digital, candidate patterns of electrical activity leading to potentiation or depression could yield transitions between these internal molecular states without necessarily inducing an associated change in synaptic efficacy.
This form of synaptic change, known as metaplasticity \cite{Abraham1996metaplasticity,Montgomery2002765}, can allow the probability of synaptic potentiation or depression to acquire a rich dependence on the history of prior changes in efficacy, thereby potentially improving memory capacity.

Theoretical studies of complex, metaplastic synapses have focused on analyzing the memory performance of a limited number of very specific molecular dynamical systems, characterized by a number of internal states in which potentiation and depression each induce a specific set of allowable transitions between states (\eg see \cref{fig:models} below).
While these models can vastly outperform simple binary synaptic switches, these analyses leave open several deep and important questions.
For example, how does the structure of a synaptic dynamical system determine its memory performance?  What are the fundamental limits of memory performance over the space of all possible synaptic dynamical systems?
What is the structural organization of synaptic dynamical systems that achieve these limits?
Moreover, from an experimental perspective, it is unlikely that all synapses can be described by a single canonical synaptic model;  just like the case of neurons, there is an incredible diversity of molecular networks underlying synapses both across species and across brain regions within a single organism \cite{Emes2012synapserev}.
In order to elucidate the functional contribution of this diverse molecular complexity to learning and memory, it is essential to move beyond the analysis of specific models and instead develop a general theory of learning and memory for complex synapses.
Moreover, such a general theory of complex synapses could aid in development of novel artificial memory storage devices.  % Sutskever, memory for learning?

Here we initiate such a general theory by proving upper bounds on the memory curve associated with any synaptic dynamical system, within the well established ideal
observer framework of \cite{Fusi2005cascade,Fusi2007multistate,Barrett2008discrete}.
Along the way we develop principles based on first passage time theory to order the structure of synaptic dynamical systems and relate this structure to memory performance.
We summarize our main results in the \hyperref[sec:disc]{discussion section}.

\section{Overall framework: synaptic models and their memory curves}\label{sec:setup}

In this section, we describe the class of models of synaptic plasticity that we are studying and how we quantify their memory performance.
In the subsequent sections, we will find upper bounds on this performance.

We use a well established formalism for the study of learning and memory with complex synapses (see \cite{Fusi2005cascade,Fusi2007multistate,Barrett2008discrete}).
In this approach, electrical patterns of activity corresponding to candidate potentiating and depressing plasticity events occur randomly and independently at all synapses at a Poisson rate $r$.  These events reflect possible synaptic changes due to either spontaneous network activity, or the storage of new memories.
We let  $f\pot$ and  $f\dep$ denote the fraction of these events that are candidate potentiating or depressing events respectively.
Furthermore, we assume our synaptic model has $M$ internal molecular functional states, and that a candidate potentiating (depotentiating) event induces a stochastic transition in the internal state described by an $M \times M$ discrete time Markov transition matrix  $\M\pot$ ($\M\dep$).
In this framework, the states of different synapses will be independent, and the entire synaptic population can be fully described by the probability distribution across these states, which we will indicate with the row-vector $\prob(t)$.
Thus the $i$'th component of $\prob(t)$ denotes the fraction of the synaptic population in state $i$.
Furthermore, each state $i$ has its own synaptic weight, $\w_i$, which we take, in the worst case scenario, to be restricted to two values.
After shifting and scaling these two values,  we can assume they are $\pm1$, without loss of generality.

We also employ an ``ideal observer'' approach to the memory readout, where the synaptic weights are read directly.
This provides an upper bound on the quality of any readout using neural activity.

For any single memory, stored at time $t=0$, we assume there will be an ideal pattern of synaptic weights across a population of $N$ synapses, the $N$-element vector $\wvi$, that is $+1$ at all synapses that experience a candidate potentiation event, and $-1$ at all synapses that experience a candidate depression event at the time of memory storage.
We assume that any pattern of synaptic weights close to $\wvi$ is sufficient to recall the memory.
However, the actual pattern of synaptic weights at some later time, $t$, will change to $\wv(t)$ due to further modifications from the storage of subsequent memories.
We can use the overlap between these, $\wvi\cdt\wv(t)$, as a measure of the quality of the memory.
As $t\to\infty$, the system will return to its steady state distribution which will be uncorrelated with the memory stored at $t=0$.
The probability distribution of the quantity $\wvi\cdt\wv(\infty)$ can be used as a ``null model'' for comparison.

The extent to which the memory has been stored is described by a signal-to-noise ratio (SNR) \cite{Fusi2005cascade,Fusi2007multistate}:
%
\begin{equation}\label{eq:SNRdef}
  \snr(t) = \frac{\av{\wv_\text{ideal}\cdt\wv(t)} - \av{\wv_\text{ideal}\cdt\wv(\infty)}}
     {\sqrt{\var(\wv_\text{ideal}\cdt\wv(\infty))}}.
\end{equation}
%
The noise in the denominator is essentially $\sqrt{N}$.
There is a correction when potentiation and depression are imbalanced, but this will not affect the upper bounds that we will discuss below and will be ignored in the subsequent formulae.

A simple average memory curve can be derived as follows.
All of the preceding plasticity events, prior to $t=0$, will put the population of synapses in its steady-state distribution, $\eq$.
The memory we are tracking at $t=0$ will change the internal state distribution to $\eq\M\pot$ (or $\eq\M\dep$) in those synapses that experience a candidate potentiation (or depression) event.
As the potentiating/depressing nature of the subsequent memories is independent of $\wvi$, we can average over all sequences, resulting in the evolution of the probability distribution:
%
\begin{multline}\label{eq:evol}
  \diff{\prob(t)}{t} = r\prob(t)\frg,
  \\ \text{where} \quad
  \frg = f\pot\M\pot + f\dep\M\dep - \I.
\end{multline}
%
Here $\frg$ is a continuous time transition matrix that models the process of forgetting the memory stored at time $t=0$ due to random candidate potentiation/depression events occurring at each synapse due to the storage of subsequent memories.
Its stationary distribution is $\eq$.

This results in the following SNR
%
\begin{equation}\label{eq:SNRcalc}
  \snr(t) = \sqrt{N}(2 f\pot f\dep) \eq \brk{\M\pot-\M\dep} \e^{rt\frg} \w.
\end{equation}
%
%The factor of $\eq$ describes the synapses being in the steady-state distribution before the memory is encoded. The factor of $(\M\pot-\M\dep)$ comes from the encoding of the memory at $t=0$, with $\wvi$ being $\pm1$ in synapses that are potentiated/depotentiated. The factor of $\e^{rt\frg}$ describes the subsequent evolution of the probability distribution, averaged over all sequences of plasticity events, and the factor of $\w$ indicates the readout via the synaptic weight.
A detailed derivation of this formula can be found in the \supp.
We will frequently refer to this function as the memory curve.
It can be thought of as the excess fraction of synapses (relative to equilibrium) that maintain their ideal synaptic strength at time $t$, as dictated by the stored memory at time $t=0$.

Much of the previous work on these types of complex synaptic models has focused on understanding the memory curves of specific models, or choices of $\M\potdep$.
Two examples of these models are shown in \cref{fig:models}. We see that they have different memory properties.
The serial model performs relatively well at one particular timescale, but it performs poorly at other times.
The cascade model does not perform quite as well at that time, but it maintains its performance over a wider range of timescales.

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \panel{\includegraphics[width=0.25\linewidth]{cascade.svg}}\label{fig:cascade}
  \panel{\includegraphics[width=0.25\linewidth]{multistate.svg}}\label{fig:multistate}
  \panel{\includegraphics[width=0.3\linewidth]{cascms.eps}}\label{fig:cascms}
 \end{myenuma}
 \end{center}
  \caption[Models of complex synapses]{Models of complex synapses.
  (\ref{fig:cascade}) The cascade model of \cite{Fusi2005cascade}, showing transitions between states of high/low synaptic weight (red/blue circles) due to potentiation/depression (solid red/dashed blue arrows).
  (\ref{fig:multistate}) The serial model of \cite{Leibold2008serial}.
  (\ref{fig:cascms}) The memory curves of these two models, showing the decay of the signal-to-noise ratio (to be defined \hyperref[sec:setup]{below}) as subsequent memories are stored.
  \label{fig:models}}
\end{figure}

In this work, rather than analyzing specific models,  we take a different approach, in order to obtain a more general theory.
We consider the \emph{entire} space of these models and find upper bounds on the memory capacity of \emph{any} of them.
The space of models with a fixed number of internal states $M$ is parameterized by the pair of $M \times M$ discrete time stochastic
transition matrices $\M\pot$ and $\M\dep$, in addition to $f\potdep$.  The parameters must satisfy the following constraints:
%
\begin{equation}\label{eq:constr}
\begin{gathered}
\begin{aligned}
  \M\potdep_{ij} &\in [0,1], &\qquad
  \sum_j \M\potdep_{ij} &= 1, \\
  f\potdep &\in [0,1],&
  f\pot + f\dep &= 1,
\end{aligned}
\\
  \eq\frg = 0, \qquad
  \sum_i \eq_i = 1, \qquad
  \w_i = \pm 1.
\end{gathered}
\end{equation}
%
The upper bounds on $\M\potdep_{ij}$ and $f\potdep$ follow automatically from the other constraints.

The critical question is: what do these constraints imply about the space of achievable memory curves in \cref{eq:SNRcalc}?
To answer this question, especially for limits on achievable memory at finite times, it will be useful to employ the eigenmode decomposition:
%
\begin{equation}\label{eq:eigendecomp}
\begin{aligned}
  \frg &= \sum_a -q_a \uv^a \vv^a,
  &\quad
  \vv^a \uv^b &= \delta_{ab},
  \\%\quad
  \frg \uv^a &= -q_a \uv^a,
  &\quad
  \vv^a \frg &= -q_a \vv^a.
\end{aligned}
\end{equation}
%
Here $q_a$ are the negative of the eigenvalues of the forgetting process $\frg$, $\uv^a$ are the right (column) eigenvectors and $\vv^a$ are the left (row) eigenvectors.
This decomposition allows us to write the memory curve as a sum of exponentials,
%
\begin{equation}\label{eq:SNReigen}
\begin{aligned}
  \snr(t) &= \sqrt{N}\sum_a \CI_a \e^{-rt/\tau_a},
%  &\quad \text{where}&\;&
%  \CI_a &= \prn{2 f\pot f\dep} \eq (\M\pot-\M\dep) \uv^a  \vv^a \w,\\&
%  & \text{and}&&
%  \tau_a &= \frac{1}{q_a}.
\end{aligned}
\end{equation}
%
where $\;\CI_a = (2 f\pot f\dep) \eq (\M\pot-\M\dep) \uv^a  \vv^a \w\;$ and $\;\tau_a = {1}/{q_a}$.
We can then ask the question: what are the constraints on these quantities, namely eigenmode initial SNR's, $\CI_a$, and time constants, $\tau_a$, implied by the constraints in \cref{eq:constr}?
We will derive some of these constraints in the next section.

\section{Upper bounds on achievable memory capacity}\label{sec:bounds}

In the previous section, in \cref{eq:SNRcalc} we have described an analytic expression for a memory curve as a function of the structure of a synaptic dynamical system, described by the pair of stochastic transition matrices  $\M\potdep$.
Since the performance measure for memory is an entire memory curve, and not just a single number, there is no universal scalar notion of optimal memory in the space of synaptic dynamical systems.
Instead there are tradeoffs between storing proximal and distal memories;  often attempts to increase memory at late (early) times by changing $\M\potdep$, incurs a performance loss in memory at early (late) times in specific models considered so far \cite{Fusi2005cascade,Fusi2007multistate,Leibold2008serial}.
Thus our end goal, achieved \hyperref[sec:env]{below}, is to derive an envelope memory curve in the SNR-time plane, or a curve that forms an upper-bound on the {\it entire} memory curve of {\it any} model.
In order to achieve this goal, in this section, we must first derive upper bounds, over the space of all possible synaptic models, on two different scalar functions of the memory curve: its initial SNR, and the area under the memory curve.
In the process of upper-bounding the area, we will develop an essential framework to organize the structure of synaptic dynamical systems based on first passage time theory.

\subsection{Bounding initial SNR}\label{sec:initial}

We now give an upper bound on the initial SNR,
%
\begin{equation}\label{eq:init}
  \snr(0) = \sqrt{N}\prn{2 f\pot f\dep} \eq \prn{\M\pot-\M\dep} \w,
\end{equation}
%
over \emph{all} possible models and also find the class of models that saturate this bound.
A useful quantity is the equilibrium probability flux between two disjoint sets of states, $\CA$ and $\CB$:
%
\begin{equation}\label{eq:flux}
  \F_{\CA\CB} = \sum_{i\in\CA} \sum_{j\in\CB} r \eq_i \frg_{ij}.
\end{equation}
%
The initial SNR is closely related to the flux from the states with $\w_i=-1$ to those with $\w_j=+1$ (see \supp):
%
\begin{equation}\label{eq:initflux}
  \snr(0) \leq \frac{4\sqrt{N}\F_{-+}}{r}\,.
\end{equation}
%
This inequality becomes an equality if potentiation never decreases the synaptic weight and depression never increases it, which should be a property of any sensible model.

To maximize this flux, potentiation from a weak state must be guaranteed to end in a strong state, and depression must do the reverse.
An example of such a model is shown in \cref{fig:max_init}(\ref{fig:max_init_pot},\ref{fig:max_init_dep}).
These models have a property known as ``lumpability'' (see \cite[\S6.3]{kemeny1960finite} for the discrete time version and \cite{burke1958markovian,Ball1993Lumpability} for continuous time).
They are completely equivalent to a two state model with transition probabilities equal to 1, as shown in \cref{fig:max_init}(\ref{fig:binary_det}).
In particular, they have the same memory curve.

This two state model has the equilibrium distribution $\eq=(f\dep,f\pot)$ and its flux is given by $\F_{-+} = rf\pot f\dep$.
This is maximized when $f\pot=f\dep=\half$, leading to the upper bound:
%
\begin{equation}\label{eq:max_init}
  \snr(0) \leq \sqrt{N}.
\end{equation}
%
We note that while this model has high initial SNR, it also has very fast memory decay -- with a timescale $\tau\sim\frac{1}{r}$.
As the synapse is very plastic, the initial memory is encoded very easily, but the subsequent memories also overwrite it rapidly.
This is one example of the tradeoff between optimizing memory at early versus late times.
\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \panel{\hp\includegraphics[width=0.24\linewidth]{max_init_pot.svg}}\label{fig:max_init_pot}\hp
  \panel{\includegraphics[width=0.24\linewidth]{max_init_dep.svg}}\label{fig:max_init_dep}\hp
  \panel{\hp\includegraphics[width=0.1\linewidth]{binary_det.svg}}\label{fig:binary_det}
  \end{myenuma}
 \end{center}
  \caption[Synaptic models that maximize initial SNR]{Synaptic models that maximize initial SNR.
  (\ref{fig:max_init_pot}) For potentiation, all transitions starting from a weak state lead to a strong state, and the probabilities for all transitions leaving a given weak state sum to 1.
  (\ref{fig:max_init_pot}) Depression is similar to potentiation, but with strong and weak interchanged.
  (\ref{fig:binary_det}) The equivalent two state model, with transition probabilities under potentiation and depression equal to one.
  \label{fig:max_init}}
\end{figure}

\subsection{Imposing order on internal states through first passage times}\label{sec:order}

Our goal is to understand the relationship between structure and function in the space of all possible synaptic models.
This goal is complicated by the fact that this space contains many different possible network topologies, encoded in the nonzero matrix elements of $\M\potdep$.
To systematically analyze this entire space, we develop an important organizing principle using the theory of first passage times in the stochastic process of forgetting, described by $\frg$.
The mean first passage time matrix, $\smash{\fptb_{ij}}$, is defined as the average time it takes to reach state $j$ for the first time, starting from state $i$.
The diagonal elements are defined to be zero.

A remarkable theorem we will exploit is that the quantity
%
\begin{equation}\label{eq:kemenyconst}
  \eta \equiv \sum_j \fptb_{ij}\eq_j,
\end{equation}
%
is independent of the starting state $i$.
It is known as Kemeny's constant \cite[see][\S4.4]{kemeny1960finite}.
Intuitively, \cref{eq:kemenyconst} states that the average time it takes to reach any state, weighted by its equilibrium probability, is independent of the starting state, implying a hidden constancy inherent in any stochastic process.

In the context of complex synapses, we can define the partial sums
%
\begin{equation}\label{eq:kemenypm}
   \eta^+_i = \sum_{j\in+} \fptb_{ij} \eq_j,
   \qquad
   \eta^-_i = \sum_{j\in-} \fptb_{ij} \eq_j.
\end{equation}
%
These can be thought of as the average time it takes to reach the strong/weak states respectively.
Using these definitions, we can then impose an order on the states by arranging them in order of decreasing $\smash{\eta^+_i}$ or increasing $\smash{\eta^-_i}$.
Because $\smash{\eta^+_i+\eta^-_i=\eta}$ is independent of $i$, the two orderings are the same.
In this order, which depends sensitively on the structure of $\M\potdep$, states later (to the right in figures below) can be considered to be more potentiated than states earlier (to the left in figures below), despite the fact that they have the same synaptic efficacy.
In essence, in this order, a state is considered to be more potentiated if the average time it takes to reach all the strong efficacy states is shorter.
We will see that synaptic models that optimize various measures of memory have an exceedingly simple structure when, and only when, their states are arranged in this order.%
\footnote{Note that we do not need to worry about the order of the $\eta^\pm_i$ changing during the optimization: necessary conditions for a maximum only require that there is no infinitesimal perturbation that increases the area. Therefore we need only consider an infinitesimal neighborhood of the model, in which the order will not change.\label{ft:etaorder}}

\subsection{Bounding area}\label{sec:area}

>>>>

@@

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\iftoggle{main} {\author[1]{Subhaneil Lahiri}
\author{Surya Ganguli}}
{\author{Lahiri \textit{et al.}}}

@@

\iftoggle{main}{%
\affil{Department of Applied Physics, Stanford University, Stanford CA}
% Please add here a significance statement to explain the relevance of your work
\significancestatement{Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality.}
% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Please provide details of author contributions here.}
\authordeclaration{Please declare any conflict of interest here.}
%\equalauthors{\textsuperscript{1}A.O.(Author One) and A.T. (Author Two) contributed equally to this work (remove if not applicable).}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: sulahiri@stanford.edu}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
\keywords{synaptic plasticity | learning | memory | neuroscience}
%\abbreviations{SNR, signal-to-noise ratio; LTP, long term potentiation;LTD, long term depression}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  An incredible gulf separates theoretical models of synapses, often described solely by a single scalar value denoting the size of a postsynaptic potential, from the immense complexity of molecular signaling pathways underlying real synapses.
  To understand the functional contribution of such molecular complexity to learning and memory, it is essential to expand our theoretical conception of a synapse from a single scalar to an entire dynamical system with many internal molecular functional states.
  Moreover, theoretical considerations alone demand such an expansion; network models with scalar synapses assuming finite numbers of distinguishable synaptic strengths have strikingly limited memory capacity.
  This raises the fundamental question, how does synaptic complexity give rise to memory?
  To address this, we develop new mathematical theorems elucidating the relationship between the structural organization and memory properties of complex synapses that are themselves molecular networks.
  Moreover, in proving such theorems, we uncover a framework, based on first passage time theory, to impose an order on the internal states of complex synaptic models, thereby simplifying the relationship between synaptic structure and function.

  Overall, we uncover general design principles governing the functional organization of complex molecular networks, and suggest new experimental observables in synaptic physiology, based on first %passage time theory, that connect molecular complexity to memory.
\end{abstract}
}{}

@@
